---
author: "Amelia Hines"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Exploratory SILVA 138.2

## Libraries
```{r}
library(ggplot2)
library(vegan)
library(stringr)
library(tidyverse)

```


## Emu Data
```{r}
emu1 <- read.table("emu_silva138.2/barcode01_unmapped.fastq.gz_silva_results_rel-abundance.tsv", 
                   sep = "\t",
                   header = TRUE)

emu_path <- list.files(path = "emu_silva138.2",
                       pattern = "\\.tsv$",
                       full.names = TRUE)

emu_abundance <- do.call(rbind, lapply(emu_path, 
                                       function(file) {
  df <- read.table(file, 
                   sep = "\t", 
                   header = TRUE)
  df$Sample_Name <- tools::file_path_sans_ext(basename(file))
  return(df)
})) %>% 
  setNames(c("taxi_id", "abundance", "superkingdom", "phylum", "class", "order", "family", "genus", "species", "estimated_counts", "sample_name")) %>% 
  mutate(sample_name = sub("_.*", "", sample_name)) %>% 
  filter(!taxi_id == "unmapped" &
           !taxi_id == "mapped_filtered" &
           !taxi_id == "mapped_unclassified")


```

```{r}
taxa_summary <- 
  emu_abundance  %>% 
  group_by(sample_name) %>% 
  summarize(taxa = n())
```

## Filter by Abundance Threshold
```{r}

emu_aggregated <- 
  emu_abundance %>% 
  select(sample_name, superkingdom, phylum, class, order, family, genus, species, abundance, estimated_counts) %>% 
  aggregate(. ~ sample_name + species + genus + family + order + class + phylum + superkingdom, sum)


emu_filtered <- 
  emu_aggregated %>% 
group_by(sample_name) %>% 
  mutate(totalreads = sum(estimated_counts),
         ra = estimated_counts/totalreads) %>% 
  filter(ra > 0.0017) %>% 
  select(-c(totalreads, ra))


sum1 <- 
  emu_filtered %>%
  group_by(sample_name) %>% 
  summarize(total_sample_reads = sum(estimated_counts))

print(sum1)

sum2 <-
    emu_abundance %>% 
    group_by(sample_name) %>% 
   summarize(total_sample_reads = sum(estimated_counts))
   print(sum2)
   
sum3 <-
  emu_aggregated %>% 
  group_by(sample_name) %>% 
  summarize(total_sample_reads = sum(estimated_counts))
print(sum3)


sum4 <-
  emu_filtered %>% 
  group_by(sample_name) %>% 
  summarize(total_sample_abundance = sum(abundance))
print(sum4)

```

## Merge and filter metadata and emu abundance
```{r}
metadata <- read.csv("Bison_nanopore_metadata (1).csv") 
   emu_abundance_merged <- merge(emu_abundance, 
                                 metadata, 
                                 by.x = "sample_name", 
                                 by.y =  "BarcodeNumber") 
   
     emu_abundance_no_controls <- filter(emu_abundance_merged, 
                                         !Type == "pos" &
                                           !Type == "neg")
     emu_abundance_no_controls
     
     emu_abundance_wbnp_samples <- filter(emu_abundance_merged, 
                                          Type == "wbnp")
     emu_abundance_wbnp_samples
     
     emu_abundance_Pos_controls <- filter(emu_abundance_merged, 
                                          Type == "pos")
     emu_abundance_Pos_controls
     
     emu_abundance_Neg_controls <- filter(emu_abundance_merged, 
                                          Type == "neg")
     emu_abundance_Neg_controls

```

## Changing Taxa Names
```{r}
emu_abundance_merged <-
  emu_abundance_merged %>%
  mutate(family = as.character(family), 
         genus = as.character(genus),
         genus = if_else(grepl("UCG-010", family, ignore.case = TRUE), "UCG-010", genus))

emu_abundance_merged <- 
  emu_abundance_merged %>%
  mutate(family = as.character(family),
         family = if_else(grepl("UCG-010", family, ignore.case = TRUE), "Ruminococcaceae", family))

emu_abundance_merged <-
  emu_abundance_merged %>%
  mutate(family = as.character(family), 
         genus = as.character(genus),
         genus = if_else(grepl("\\[Eubacterium\\]_coprostanoligenes_group", family, ignore.case = TRUE), "[Eubacterium]_coprostanoligenes_group", genus))

emu_abundance_merged <- 
  emu_abundance_merged %>%
  mutate(family = as.character(family),
         family = if_else(grepl("\\[Eubacterium\\]_coprostanoligenes_group", family, ignore.case = TRUE), "Oscillospiraceae", family))

emu_abundance_merged <-
  emu_abundance_merged %>%
  mutate(family = as.character(family), 
         genus = as.character(genus),
         genus = if_else(grepl("UCG-011", family, ignore.case = TRUE), "UCG-011", genus))

emu_abundance_merged <- 
  emu_abundance_merged %>%
  mutate(family = as.character(family),
         family = if_else(grepl("UCG-011", family, ignore.case = TRUE), "Ruminococcaceae", family))
```

## Plot Controls

### Plot Positives
```{r}
# Plot of phylum abundance
superkingdom_abundance_Pos_controls <- 
  emu_abundance_Pos_controls %>% 
  select(superkingdom, 
         sample_name, 
         estimated_counts) %>% 
  aggregate(. ~ superkingdom + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = superkingdom)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+                scale_y_continuous(expand = c(0,0), 
                                                                                        labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560","#A1527F", "#CC79A7"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Sample SuperkingdomAbundance",
         fill = "Superkingdom")
superkingdom_abundance_Pos_controls

# Plot of phylum abundance
phylum_abundance_Pos_controls <- 
  emu_abundance_Pos_controls %>% 
  select(phylum, 
         sample_name, 
         estimated_counts) %>% 
  aggregate(. ~ phylum + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+                scale_y_continuous(expand = c(0,0), 
                                                                                        labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560","#A1527F", "#CC79A7"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Sample Phylum Abundance",
         fill = "Phylum")
phylum_abundance_Pos_controls
ggsave("Figures/pos_control_phylum_abundance_silva.jpg",
       phylum_abundance_Pos_controls, 
       width = 8, 
       height = 7, 
       units = "in")


# Plot of family abundance
family_abundance_Pos_controls <- 
  emu_abundance_Pos_controls %>% 
  select(family, 
         sample_name, 
         estimated_counts) %>% 
  aggregate(. ~ family + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = family)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+                scale_y_continuous(expand = c(0,0), 
                   labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#4E7705", "#6D9F06", "#97CE2F", "#BDEC6F", "#DDFFA0", "#FFD5AF","#FCB076", "#F09163"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Sample Family Abundance",
         fill = "Family")
family_abundance_Pos_controls

```

## Plot Species Abundance
```{r}
ra_fun <- function(x) {
  x/sum(x)
}
# Convert long to wide dataset
wide_emu_pos_control_species <- 
  emu_abundance_merged %>% 
  filter(Type == "pos") %>% 
  select(species, sample_name, estimated_counts) %>% 
  aggregate(. ~ species + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

species_abundance_Pos_controls <-
  wide_emu_pos_control_species %>% 
  select(species, all_of(emu_abundance_Pos_controls$sample_name)) %>%
  aggregate(. ~ species, sum) %>% 
  mutate(TotalReads = rowSums(across(emu_abundance_Pos_controls$sample_name))) %>% 
  arrange(., desc(TotalReads)) %>% 

  mutate(Group = ifelse(row_number() < 16, species, "Other")) %>% 

  select(!TotalReads) %>% 

  pivot_longer(!c(species, Group), names_to = "Sample", values_to = "Reads") %>% 
  group_by(Group, Sample) %>% 
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Sample, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Species Abundance",
         fill = "Species")
species_abundance_Pos_controls
ggsave("Figures/pos_controls_species_abundance_silva.jpg", species_abundance_Pos_controls, 
       width = 8, 
       height = 7, 
       units = "in")
```

### Plot Negatives
```{r}
# Plot of phylum abundance
phylum_abundance_Neg_controls <- 
  emu_abundance_Neg_controls %>% 
  select(phylum, sample_name, estimated_counts) %>% 
  aggregate(. ~ phylum + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5),
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
  scale_y_continuous(expand = c(0,0), 
                     labels = scales::percent)+
    scale_fill_manual(values = c("#7D3560","#A1527F", "#CC79A7", "#E794C1", "#EFB6D6","#E0F1FF", "#B0DAF9"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Negative Control Phylum Abundance",
         fill = "Phylum")
phylum_abundance_Neg_controls
ggsave("Figures/neg_controls_phylum_abundance_silva.jpg", phylum_abundance_Neg_controls, 
       width = 8, 
       height = 7, 
       units = "in")

```


## Plot WBNP Samples

### Plot Phylum Abundance
```{r}
phylum_abundance_samples <- 
  emu_abundance_wbnp_samples %>% 
  select(phylum, sample_name, estimated_counts, Herd) %>% 
  aggregate(. ~ phylum + sample_name + Herd, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 5))+
  facet_wrap(. ~ Herd, 
             nrow = 1, 
             scales = "free_x")+
  scale_y_continuous(expand = c(0,0), 
                     labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "WBNP Samples Phylum by Herd",
         fill = "Phylum")
phylum_abundance_samples
ggsave("Figures/wbnp_phylum_abundance_byherd_silva.jpg", 
       phylum_abundance_samples, 
       width = 14, 
       height = 7, 
       units = "in")

```

### Plot Class Abundance
```{r}

class_abundance_samples <- 
  emu_abundance_wbnp_samples %>% 
  select(class, sample_name, estimated_counts, Herd) %>% 
  aggregate(. ~ class + sample_name + Herd, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = class)) +
    geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 5))+
  facet_wrap(. ~ Herd, 
             nrow = 1, 
             scales = "free_x")+
  scale_y_continuous(expand = c(0,0), 
                     labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705", "#9D654C", "#C17754","#F09163", "#FCB076","#FFD5AF"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "WBNP Samples Class by Herd",
         fill = "Class")
class_abundance_samples
```

### Plot Family Abundance

```{r}

ra_fun <- function(x) {
  x/sum(x)
}
# Convert long to wide dataset
wide_emu_family <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(family, sample_name, estimated_counts) %>% 
  aggregate(. ~ family + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

family_abundance_samples <-
  wide_emu_family %>% 
# select taxonomic level of interest and all samples (that are each columns)
  select(family, all_of(emu_abundance_wbnp_samples$sample_name)) %>%
# aggregate based  on taxonomic level of interest
  aggregate(. ~ family, sum) %>% 
# create column of sum of reads per each family
  mutate(TotalReads = rowSums(across(emu_abundance_wbnp_samples$sample_name))) %>% 
# sort dataframe based on total reads in descending order (max at top)
  arrange(., desc(TotalReads)) %>% 
# create a group column, pick the top 16 families with the most reads and label them with Family name, name all others "Other"
  mutate(Group = ifelse(row_number() < 16, family, "Other")) %>% 
# get rid of total reads column, unneeded
  select(!TotalReads) %>% 
# pivot to long format, keeping Family and Group columns separate
  pivot_longer(!c(family, Group), names_to = "Sample", values_to = "Reads") %>% 
# Group by new Group variable and Sample
  group_by(Group, Sample) %>% 
# create variable to sum reads for each Group (16 families + Other)
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Sample, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Family Abundance by Sample",
         fill = "Family")
family_abundance_samples
ggsave("Figures/wbnp_family_abundance_silva.jpg", 
       family_abundance_samples, 
       width = 14, 
       height = 7, 
       units = "in")
```



## Plot by Herd

### Plot Phylum Abundance
```{r}

# Plot of phylum abundance
phylum_abundance_herd <- 
  emu_abundance_wbnp_samples %>% 
  select(phylum, Herd, estimated_counts) %>% 
  aggregate(. ~ phylum + Herd, sum) %>% 
    ggplot(., aes(x = Herd, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705"))+
    labs(x = "Herd", 
         y = "Proportion of Counts",
         title = "Phylum Abundance By Herd",
         fill = "Phylum")
phylum_abundance_herd
ggsave("Figures/herd_phylum_abundance_silva.jpg", 
       phylum_abundance_herd, 
       width = 8, 
       height = 7, 
       units = "in")

```

### Plot Family Abundance
```{r}
#Plot of family abundance
ra_fun <- function(x) {
  x/sum(x)
}
# Convert long to wide dataset
wide_emu_family_herd <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(family, Herd, estimated_counts) %>% 
  aggregate(. ~ family + Herd, sum) %>% 
pivot_wider(names_from = "Herd", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

family_abundance_herd <-
  wide_emu_family_herd %>% 
# select taxonomic level of interest and all samples (that are each columns)
  select(family, all_of(emu_abundance_wbnp_samples$Herd)) %>%
# aggregate based  on taxonomic level of interest
  aggregate(. ~ family, sum) %>% 
# create column of sum of reads per each family
  mutate(TotalReads = rowSums(across(emu_abundance_wbnp_samples$Herd))) %>% 
# sort dataframe based on total reads in descending order (max at top)
  arrange(., desc(TotalReads)) %>% 
# create a group column, pick the top 16 families with the most reads and label them with Family name, name all others "Other"
  mutate(Group = ifelse(row_number() < 16, family, "Other")) %>% 
# get rid of total reads column, unneeded
  select(!TotalReads) %>% 
# pivot to long format, keeping Family and Group columns separate
  pivot_longer(!c(family, Group), names_to = "Herd", values_to = "Reads") %>% 
# Group by new Group variable and Sample
  group_by(Group, Herd) %>% 
# create variable to sum reads for each Group (16 families + Other)
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Herd, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Herd", 
         y = "Proportion of Counts",
         title = "Family Abundance by Herd",
         fill = "Family")
family_abundance_herd
ggsave("Figures/herd_family_abundance_silva.jpg",
       family_abundance_herd,
       width = 8,
       height = 7,
       units = "in")

```


## Plot Reads

### Plot Total Reads
```{r}
new_labels <- c("panp" = "PANP Samples", 
                "neg" = "Negative Controls", 
                "pos" = "Positive Controls", 
                "wbnp" = "WBNP Samples")
total_reads <- 
  emu_abundance_merged %>% 
  group_by(sample_name) %>% 
  summarize(totalreads = sum (estimated_counts)) %>% 
  merge(., metadata, 
        by.x = "sample_name", 
        by.y = "BarcodeNumber") %>% 
  ggplot(., aes(x = sample_name, 
                y = totalreads))+
  geom_bar(stat = "identity", 
           fill = "#7D3560")+
  facet_wrap(~ Type, 
             nrow = 1, 
             scales = "free_x", 
             labeller = as_labeller(new_labels))+
  scale_y_continuous(expand = c(0,0))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 5))+
  labs(x = "Sample", 
       y = "Total Reads", 
       title = "Sample Total Reads")  
total_reads
ggsave("Figures/16S_totalreads_silva.jpg", 
       total_reads, 
       width = 14, 
       height = 7, 
       units = "in")

```


## PCoA Plotting

### Distance Matrix
```{r}

ra_fun <- function(x) {
  x/sum(x)
}
wide_emu_genus <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(genus, sample_name, estimated_counts) %>% 
  aggregate(. ~ genus + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

colSums(wide_emu_genus[, 2:54])

dist_emu <-
  wide_emu_genus %>% 
  column_to_rownames("genus") %>% 
  t() %>% 
  data.frame() %>% 
  vegdist(., method = "bray")

```

### Create PCoA
```{r}
pcoa_genus <- 
  cmdscale(dist_emu, eig = TRUE, add = TRUE)

pcoa_genus_data <-
  as.data.frame(pcoa_genus$points) %>% 
  setNames(c("PCo1", "PCo2")) %>% 
  rownames_to_column(var = "Sample") %>% 
  merge(., metadata, by.x = "Sample", by.y = "BarcodeNumber") %>% 
  mutate(eig = pcoa_genus$eig, 
         perc = round((eig/sum(eig))*100, 2))

pcoa_genus_plot <- 
  ggplot(pcoa_genus_data, aes(x = PCo1, y = PCo2, color = Herd)) +
  geom_point(size = 2) +
  stat_ellipse() +
  scale_color_manual(values = c("#E794C1", "#73B8EF","#BDEC6F", "#FCB076", "#B7B7B7", "#6a51a3", "#a6611a", "#43BA8F", "#d7301f"))+
  labs(x = paste("PCo1 ", pcoa_genus_data$perc[1], "%"),
       y = paste("PCo2 ", pcoa_genus_data$perc[2], "%"),
       title = "Herd PCoA Plot",
       color = "Herd")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))
pcoa_genus_plot
ggsave("Figures/herd_pcoa_silva.jpg", pcoa_genus_plot, width = 14, height = 7, units = "in")

```


## Total Reads per Taxa Level

### Determining Known and Unknown
```{r}

wide_emu_order <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(order, sample_name, estimated_counts) %>% 
  aggregate(. ~ order + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

unclass_order <-
  wide_emu_order %>% 
  select(order, all_of(emu_abundance_wbnp_samples$sample_name)) %>% 
  aggregate(. ~ order, sum) %>% 
  mutate(status = if_else(order %in% c("Incertae_Sedis" ), "Unknown", "Known")) %>% 
  select(!order) %>% 
  aggregate(. ~ status, sum) %>% 
  column_to_rownames("status") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column("sample_name") %>% 
  pivot_longer(!sample_name, names_to = "Group", values_to = "Reads") %>% 
  mutate(Taxon = "Order")
  

unclass_family <-
  wide_emu_family %>% 
  select(family, all_of(emu_abundance_wbnp_samples$sample_name)) %>% 
  aggregate(. ~ family, sum) %>% 
  mutate(status = if_else(family %in% c("Coriobacteriales_Incertae_Sedis", "Incertae_Sedis" ), "Unknown", "Known")) %>% 
  select(!family) %>% 
  aggregate(. ~ status, sum) %>% 
  column_to_rownames("status") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column("sample_name") %>% 
  pivot_longer(!sample_name, names_to = "Group", values_to = "Reads") %>% 
  mutate(Taxon = "Family")

unclass_genus <-
  wide_emu_genus %>% 
  select(genus, all_of(emu_abundance_wbnp_samples$sample_name)) %>% 
  aggregate(. ~ genus, sum) %>% 
  mutate(status = if_else(genus %in% c("Incertae_Sedis" ), "Unknown", "Known")) %>% 
  select(!genus) %>% 
  aggregate(. ~ status, sum) %>% 
  column_to_rownames("status") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column("sample_name") %>% 
  pivot_longer(!sample_name, names_to = "Group", values_to = "Reads") %>% 
  mutate(Taxon = "Genus")

wide_emu_species <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(species, sample_name, estimated_counts) %>% 
  aggregate(. ~ species + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

  unclass_species <-
  wide_emu_species %>% 
  select(species, all_of(emu_abundance_wbnp_samples$sample_name)) %>% 
  aggregate(. ~ species, sum) %>% 
  mutate(status = ifelse(species %in% c("gut_metagenome", "human_gut", "bactrium_enrichment", "metagenome", "bacterium", "iron-reducing_bacterium", "unidentified_rumen", "rumen_bacterium", "unidentified", "wallaby_gut") | grepl("uncultured", species) | grepl("_sp.", species), "Unknown", "Known")) %>% 
  select(!species) %>% 
  aggregate(. ~ status, sum) %>% 
  column_to_rownames("status") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column("sample_name") %>% 
  pivot_longer(!sample_name, names_to = "Group", values_to = "Reads") %>% 
  mutate(Taxon = "Species")

```

### Plotting
```{r}
taxa_level_unclass <-
  do.call("rbind", list(unclass_species, unclass_genus, unclass_family, unclass_order)) %>% 
  mutate(Taxon = factor(Taxon, levels = c("Species", "Genus", "Family", "Order")))

taxa_level_unclass_plot <-
taxa_level_unclass %>% 
  ggplot(., aes(x = Taxon, y = Reads, fill = Group)) +
  geom_bar(stat = "identity", position = "fill") + 
  scale_fill_manual(values = c("gray80", "gray30")) +
  scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  theme_bw() +
  theme(panel.grid = element_blank(),
        text = element_text(size = 20))+
  labs(x = "Taxon",
       y = "Reads",
       title = "Classified Reads Per Taxon")
taxa_level_unclass_plot

ggsave("Figures/classified_reads_per_taxa_silva.jpg",
       taxa_level_unclass_plot,
       width = 8,
       height = 7,
       units = "in")
```


