---
author: "Amelia Hines"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Exploratory GTDB

## Libraries
```{r}
library(ggplot2)
library(tidyverse)
library(vegan)
library(stringr)

```

## Emu Data
```{r}
emu_gtdb_1 <- read.table("emu_gtdb/barcode01_unmapped.fastq.gz_gtdb_results_rel-abundance.tsv", 
                   sep = "\t",
                   header = TRUE)

emu_gtdb_path <- list.files(path = "emu_gtdb",
                       pattern = "\\.tsv$",
                       full.names = TRUE)

emu_gtdb_abundance <- do.call(rbind, lapply(emu_gtdb_path, 
                                       function(file) {
  df <- read.table(file, 
                   sep = "\t", 
                   header = TRUE)
  df$Sample_Name <- tools::file_path_sans_ext(basename(file))
  return(df)
})) %>% 
  setNames(c("taxi_id", "abundance", "species", "genus", "family", "order", "class", "phylum", "clade", "superkingdom", "subspecies", "species.subgroup", "species.group", "estimated_counts", "sample_name")) %>% 
  mutate(sample_name = sub("_.*", "", sample_name)) %>% 
  filter(!taxi_id == "unmapped" &
           !taxi_id == "mapped_filtered" &
           !taxi_id == "mapped_unclassified") 
  emu_gtdb_abundance$species.subgroup <-NULL
  emu_gtdb_abundance$subspecies <-NULL 
  emu_gtdb_abundance$species.group <-NULL 



```

```{r}
taxa_summary <- 
  emu_gtdb_abundance  %>% 
  group_by(sample_name) %>% 
  summarize(taxa = n())
```



## Filter by Abundance Threshold
```{r}

emu_gtdb_aggregated <- 
  emu_gtdb_abundance %>% 
  select(sample_name, superkingdom, phylum, class, order, family, genus, species, abundance, estimated_counts) %>% 
  aggregate(. ~ sample_name + species + genus + family + order + class + phylum + superkingdom, sum)


gtdb_filtered <- 
  emu_gtdb_aggregated %>% 
group_by(sample_name) %>% 
  mutate(totalreads = sum(estimated_counts),
         ra = estimated_counts/totalreads) %>% 
  filter(ra > 0.0017) %>% 
  select(-c(totalreads, ra))


sum1 <- 
  gtdb_filtered %>%
  group_by(sample_name) %>% 
  summarize(total_sample_reads = sum(estimated_counts))

print(sum1)

sum2 <-
    emu_gtdb_abundance_merged %>% 
    group_by(sample_name) %>% 
   summarize(total_sample_reads = sum(estimated_counts))
   print(sum2)
   
sum3 <-
  emu_gtdb_aggregated %>% 
  group_by(sample_name) %>% 
  summarize(total_sample_reads = sum(estimated_counts))
print(sum3)


sum4 <-
  gtdb_filtered %>% 
  group_by(sample_name) %>% 
  summarize(total_sample_abundance = sum(abundance))
print(sum4)

```

## Read in Metadata
```{r}
metadata <- read.csv("Bison_nanopore_metadata (1).csv")
   emu_gtdb_abundance_merged <- 
     merge(gtdb_filtered, metadata, 
         by.x = "sample_name", 
         by.y =  "BarcodeNumber", 
         all.x = TRUE) 
   
    gtdb_no_controls <- filter(emu_gtdb_abundance_merged, 
                                         !Type == "pos" &
                                           !Type == "neg")

     
     gtdb_wbnp_samples <- filter(emu_gtdb_abundance_merged, 
                                          Type == "wbnp")
    
     
     gtdb_pos_controls <- filter(emu_gtdb_abundance_merged, 
                                          Type == "pos")
     
     
     gtdb_neg_controls <- filter(emu_gtdb_abundance_merged, 
                                          Type == "neg")
     
```


## Plot Total Reads
```{r}
new_labels <- c("panp" = "PANP Samples", 
                "neg" = "Negative Controls", 
                "pos" = "Positive Controls", 
                "wbnp" = "WBNP Samples")
total_reads <- 
  emu_gtdb_abundance_merged %>% 
  group_by(sample_name) %>% 
  summarize(totalreads = sum (estimated_counts)) %>% 
  merge(., metadata, 
        by.x = "sample_name", 
        by.y = "BarcodeNumber") %>% 
  ggplot(., aes(x = sample_name, 
                y = totalreads))+
  geom_bar(stat = "identity", 
           fill = "#7D3560")+
  facet_wrap(~ Type, 
             nrow = 1, 
             scales = "free_x", 
             labeller = as_labeller(new_labels))+
  scale_y_continuous(expand = c(0,0))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 5))+
  labs(x = "Sample", 
       y = "Total Reads", 
       title = "Sample Total Reads")  
total_reads


```

## Plot Positive Controls

### Plot Phylum Abundance
```{r}
phylum_gtdb_abundance_pos_controls <- 
  gtdb_pos_controls %>% 
  select(phylum, 
         sample_name, 
         estimated_counts) %>% 
  aggregate(. ~ phylum + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+                scale_y_continuous(expand = c(0,0), 
                   labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560","#A1527F", "#CC79A7","#E794C1"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Sample Phylum Abundance",
         fill = "Phylum")
phylum_gtdb_abundance_pos_controls
ggsave("Figures/pos_control_phylum_abundance_gtdb.jpg",
       phylum_gtdb_abundance_pos_controls, 
       width = 8, 
       height = 7, 
       units = "in")
```

### Plot Family Abundance
```{r}

family_gtdb_abundance_pos_controls <- 
  gtdb_pos_controls %>% 
  select(family, 
         sample_name, 
         estimated_counts) %>% 
  aggregate(. ~ family + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = family)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+                scale_y_continuous(expand = c(0,0), 
                   labels = scales::percent)+
   scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705", "#9D654C", "#C17754","#F09163", "#FCB076","#FFD5AF","#F5F5F5", "#D6D6D6", "#B7B7B7", "#8B8B8B", "#616161"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Sample Family Abundance",
         fill = "Family")
family_gtdb_abundance_pos_controls

```
### Plot Genus Abundance
```{r}
ra_fun <- function(x) {
  x/sum(x)
}

wide_gtdb_pos_control_genus <- 
  emu_gtdb_abundance_merged %>% 
  filter(Type == "pos") %>% 
  select(genus, sample_name, estimated_counts) %>% 
  aggregate(. ~ genus + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

genus_gtdb_abundance_pos_controls <-
  wide_gtdb_pos_control_genus %>% 
  select(genus, all_of(gtdb_pos_controls$sample_name)) %>%
  aggregate(. ~ genus, sum) %>% 
  mutate(TotalReads = rowSums(across(gtdb_pos_controls$sample_name))) %>% 
  arrange(., desc(TotalReads)) %>% 

  mutate(Group = ifelse(row_number() < 16, genus, "Other")) %>% 

  select(!TotalReads) %>% 

  pivot_longer(!c(genus, Group), names_to = "Sample", values_to = "Reads") %>% 
  group_by(Group, Sample) %>% 
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Sample, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Genus Abundance",
         fill = "Genus")
genus_gtdb_abundance_pos_controls

```

### Plot Species Abundance 
```{r}

wide_gtdb_pos_control_species <- 
  emu_gtdb_abundance_merged %>% 
  filter(Type == "pos") %>% 
  select(species, sample_name, estimated_counts) %>% 
  aggregate(. ~ species + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

species_gtdb_abundance_pos_controls <-
  wide_gtdb_pos_control_species %>% 
  select(species, all_of(gtdb_pos_controls$sample_name)) %>%
  aggregate(. ~ species, sum) %>% 
  mutate(TotalReads = rowSums(across(gtdb_pos_controls$sample_name))) %>% 
  arrange(., desc(TotalReads)) %>% 

  mutate(Group = ifelse(row_number() < 14, species, "Other")) %>% 

  select(!TotalReads) %>% 

  pivot_longer(!c(species, Group), names_to = "Sample", values_to = "Reads") %>% 
  group_by(Group, Sample) %>% 
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Sample, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Positive Control Species Abundance",
         fill = "Species")
species_gtdb_abundance_pos_controls
ggsave("Figures/pos_controls_species_abundance_gtdb.jpg", species_gtdb_abundance_pos_controls, 
       width = 8, 
       height = 7, 
       units = "in")
```



## Plot Negative Controls

### Plot Phylum Abundance
```{r}
phylum_gtdb_abundance_neg_controls <- 
  gtdb_neg_controls %>% 
  select(phylum, sample_name, estimated_counts) %>% 
  aggregate(. ~ phylum + sample_name, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5),
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
  scale_y_continuous(expand = c(0,0), 
                     labels = scales::percent)+
    scale_fill_manual(values = c("#7D3560","#A1527F", "#CC79A7", "#E794C1", "#EFB6D6","#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Negative Control Phylum Abundance",
         fill = "Phylum")
phylum_gtdb_abundance_neg_controls
ggsave("Figures/neg_controls_phylum_abundance_gtdb.jpg", phylum_gtdb_abundance_neg_controls, 
       width = 8, 
       height = 7, 
       units = "in")
```

## Plot WBNP Samples

### Plot Phylum Abundance
```{r}
phylum_gtdb_abundance_samples <- 
 gtdb_wbnp_samples %>% 
  select(phylum, sample_name, estimated_counts, Herd) %>% 
  aggregate(. ~ phylum + sample_name + Herd, sum) %>% 
    ggplot(., aes(x = sample_name, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 5))+
  facet_wrap(. ~ Herd, 
             nrow = 1, 
             scales = "free_x")+
  scale_y_continuous(expand = c(0,0), 
                     labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705","#9D654C", "#C17754"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "WBNP Samples Phylum by Herd",
         fill = "Phylum")
phylum_gtdb_abundance_samples
ggsave("Figures/wbnp_phylum_abundance_byherd_gtdb.jpg", 
       phylum_gtdb_abundance_samples, 
       width = 14, 
       height = 7, 
       units = "in")
```


### Plot Family Abundance
```{r}

wide_emu_gtdb_family <- 
  emu_gtdb_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(family, sample_name, estimated_counts) %>% 
  aggregate(. ~ family + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

family_gtdb_abundance_samples <-
  wide_emu_gtdb_family %>% 
  select(family, all_of(emu_abundance_wbnp_samples$sample_name)) %>%
  aggregate(. ~ family, sum) %>% 
  mutate(TotalReads = rowSums(across(emu_abundance_wbnp_samples$sample_name))) %>% 
  arrange(., desc(TotalReads)) %>% 
  mutate(Group = ifelse(row_number() < 16, family, "Other")) %>% 
  select(!TotalReads) %>% 
  pivot_longer(!c(family, Group), names_to = "Sample", values_to = "Reads") %>% 
  group_by(Group, Sample) %>% 
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Sample, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Sample", 
         y = "Proportion of Counts",
         title = "Family Abundance by Sample",
         fill = "Family")
family_gtdb_abundance_samples
ggsave("Figures/wbnp_family_abundance_gtdb.jpg", 
       family_gtdb_abundance_samples, 
       width = 14, 
       height = 7, 
       units = "in")
```

## Plot by Herd

### Plot Phylum Abundance
```{r}

phylum_gtdb_abundance_herd <- 
  gtdb_wbnp_samples %>% 
  select(phylum, Herd, estimated_counts) %>% 
  aggregate(. ~ phylum + Herd, sum) %>% 
    ggplot(., aes(x = Herd, 
                  y = estimated_counts, 
                  fill = phylum)) +
    geom_bar(stat = "identity", 
             position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0", "#BDEC6F", "#97CE2F",  "#6D9F06", "#4E7705","#9D654C", "#C17754"))+
    labs(x = "Herd", 
         y = "Proportion of Counts",
         title = "Phylum Abundance By Herd",
         fill = "Phylum")
phylum_gtdb_abundance_herd
ggsave("Figures/herd_phylum_abundance_gtdb.jpg", 
       phylum_gtdb_abundance_herd, 
       width = 8, 
       height = 7, 
       units = "in")

```

### Plot Family Abundance
```{r}
wide_gtdb_family_herd <- 
  emu_gtdb_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(family, Herd, estimated_counts) %>% 
  aggregate(. ~ family + Herd, sum) %>% 
pivot_wider(names_from = "Herd", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

family_gtdb_abundance_herd <-
  wide_gtdb_family_herd %>% 
  select(family, all_of(emu_abundance_wbnp_samples$Herd)) %>%
  aggregate(. ~ family, sum) %>% 
  mutate(TotalReads = rowSums(across(emu_abundance_wbnp_samples$Herd))) %>% 
  arrange(., desc(TotalReads)) %>% 
  mutate(Group = ifelse(row_number() < 16, family, "Other")) %>% 
  select(!TotalReads) %>% 
  pivot_longer(!c(family, Group), names_to = "Herd", values_to = "Reads") %>% 
  group_by(Group, Herd) %>% 
  summarize(PlotReads = sum(Reads), .groups = "drop") %>%
  data.frame()  %>%
  mutate(Group = factor(Group),
         Group = forcats::fct_relevel(Group, "Other", after = Inf)) %>%
ggplot (., aes(x = Herd, 
               y = PlotReads, 
               fill = Group))+
          geom_bar(stat = "identity", position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust =1), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x.bottom = element_blank(), 
        axis.text.x.bottom = element_text(size = 8))+
   scale_y_continuous(expand = c(0,0), 
                      labels = scales::percent)+
  scale_fill_manual(values = c("#7D3560", "#A1527F", "#CC79A7", "#E794C1", "#EFB6D6", "#E0F1FF", "#B0DAF9", "#73B8EF", "#2382DA","#084C8A", "#148F77", "#009E73", "#43BA8F", "#48C9B0", "#A3E4D7", "#DDFFA0"))+
    labs(x = "Herd", 
         y = "Proportion of Counts",
         title = "Family Abundance by Herd",
         fill = "Family")
family_gtdb_abundance_herd
ggsave("Figures/herd_family_abundance_gtdb.jpg",
       family_gtdb_abundance_herd,
       width = 8,
       height = 7,
       units = "in")

```

## PCoA Plotting

### Distance Matrix
```{r}

wide_gtdb_genus <- 
  emu_abundance_merged %>% 
  filter(Type == "wbnp") %>% 
  select(genus, sample_name, estimated_counts) %>% 
  aggregate(. ~ genus + sample_name, sum) %>% 
pivot_wider(names_from = "sample_name", values_from = "estimated_counts") %>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>% 
  mutate(across(where(is.numeric), ra_fun))

colSums(wide_gtdb_genus[, 2:54])

dist_gtdb <-
  wide_gtdb_genus %>% 
  column_to_rownames("genus") %>% 
  t() %>% 
  data.frame() %>% 
  vegdist(., method = "bray")

```

### Create PCoA
```{r}
pcoa_genus_gtdb <- 
  cmdscale(dist_gtdb, eig = TRUE, add = TRUE)

pcoa_genus_gtdb_data <-
  as.data.frame(pcoa_genus_gtdb$points) %>% 
  setNames(c("PCo1", "PCo2")) %>% 
  rownames_to_column(var = "Sample") %>% 
  merge(., metadata, by.x = "Sample", by.y = "BarcodeNumber") %>% 
  mutate(eig = pcoa_genus_gtdb$eig, 
         perc = round((eig/sum(eig))*100, 2))

pcoa_genus_gtdb_plot <- 
  ggplot(pcoa_genus_gtdb_data, aes(x = PCo1, y = PCo2, color = Herd)) +
  geom_point(size = 2) +
  stat_ellipse() +
  scale_color_manual(values = c("#E794C1", "#73B8EF","#BDEC6F", "#FCB076", "#B7B7B7", "#6a51a3", "#a6611a", "#43BA8F", "#d7301f"))+
  labs(x = paste("PCo1 ", pcoa_genus_gtdb_data$perc[1], "%"),
       y = paste("PCo2 ", pcoa_genus_gtdb_data$perc[2], "%"),
       title = "Herd PCoA Plot",
       color = "Herd")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))
pcoa_genus_gtdb_plot
ggsave("Figures/herd_pcoa_gtdb.jpg", pcoa_genus_gtdb_plot, width = 14, height = 7, units = "in")

```
